---
title: "CPI Inflation Distribution Ridgeline Plot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Setup

This notebook fetches live CPI data from the BLS public API and renders a trimmed distribution ridgeline plot. No API key required.

```{r install, eval=FALSE}
# Run this cell only once to install packages
install.packages(c(
  "tidyverse", "httr", "data.table", "magrittr",
  "ggridges", "lubridate", "janitor", "viridis", "scales"
))
```

```{r libraries}
library(tidyverse)
library(httr)
library(data.table)
library(magrittr)
library(ggridges)
library(lubridate)
library(janitor)
library(viridis)
library(scales)
```

---

## Parameters (edit these)

```{r params}
# How far back to show on the chart (must be YYYY-MM-DD format)
chart_start_date <- "2018-06-01"

# Trim percentage (e.g. 0.15 trims bottom 15% and top 15%)
trim_lower <- 0.15
trim_upper <- 0.85

# How many quarters back to show (e.g. 4 = last 4 quarterly observations)
n_quarters <- 10
```

---

## Load CPI Data from BLS

```{r load_data}
user_email <- "rortybomb@gmail.com"  # BLS requests a contact email in the user-agent

cpi_data <- GET(
  "https://download.bls.gov/pub/time.series/cu/cu.data.0.Current",
  user_agent(user_email)
) %>%
  content(as = "text") %>%
  fread()

cpi_data <- cpi_data %>%
  clean_names() %>%
  mutate(
    value = as.numeric(value),
    date  = as.Date(
      paste(substr(period, 2, 3), "01", substr(year, 3, 4), sep = "/"),
      "%m/%d/%y"
    )
  )
```

## Load Series & Item Metadata

```{r load_series}
series <- GET(
  "https://download.bls.gov/pub/time.series/cu/cu.series",
  user_agent(user_email)
) %>%
  content(as = "text") %>%
  fread() %>%
  clean_names()

items <- GET(
  "https://download.bls.gov/pub/time.series/cu/cu.item",
  user_agent(user_email)
) %>%
  content(as = "text") %>%
  fread()

series <- inner_join(series, items, by = "item_code")

cpi_data <- cpi_data %>%
  mutate(series_id = trimws(series_id)) %>%
  inner_join(series, by = "series_id")
```

## Load Weights from BLS (no local files needed)

Binder has no local filesystem for your CSVs, so weights are fetched directly from the BLS supplemental files.

```{r load_weights}
# Primary weights
cpi_weights_url <- "https://download.bls.gov/pub/time.series/cu/cu.weights"

cpi_weights_raw <- GET(cpi_weights_url, user_agent(user_email)) %>%
  content(as = "text") %>%
  fread() %>%
  clean_names()

# The BLS weights file has: series_id, year, weight
# Join by series_id + year to attach weights
cpi_weights_joined <- cpi_weights_raw %>%
  mutate(series_id = trimws(series_id))

cpi_data <- left_join(cpi_data, cpi_weights_joined, by = c("series_id", "year"))
```

## Load Median CPI Components

The Cleveland Fed's median CPI component list is replicated inline so no local file is needed.

```{r median_components}
# Standard Cleveland Fed median CPI components (as of 2024 weights revision)
# Update this vector if the Fed revises the component list
median_component_names <- c(
  "Airline fares",
  "Alcoholic beverages at home",
  "Alcoholic beverages away from home",
  "Apparel",
  "Car and truck rental",
  "Education",
  "Food away from home",
  "Food at home",
  "Fuels and utilities",
  "Household furnishings and operations",
  "Household operations",
  "Housekeeping supplies",
  "Infants' and toddlers' apparel",
  "Intracity transportation",
  "Jewelry and watches",
  "Medical care commodities",
  "Medical care services",
  "Men's and boys' apparel",
  "Motor vehicle insurance",
  "Motor vehicle maintenance and repair",
  "New vehicles",
  "Other goods",
  "Other personal services",
  "Other services",
  "Personal care",
  "Pet services including veterinary",
  "Public transportation",
  "Recreation",
  "Rent of primary residence",
  "Shelter",
  "Tobacco and smoking products",
  "Used cars and trucks",
  "Video and audio",
  "Water and sewer and trash collection services",
  "Women's and girls' apparel"
)
```

---

## Compute Trimmed Distribution

```{r compute_median}
start_month <- month(max(cpi_data$date, na.rm = TRUE))
quarters    <- ((seq(start_month, start_month + n_quarters - 1, by = 1) - 1) %% 12) + 1

median_df <- cpi_data %>%
  filter(
    item_name %in% median_component_names |
    item_name == "Owners' equivalent rent of residences"
  ) %>%
  filter(period != "M13", seasonal == "S") %>%
  arrange(date) %>%
  group_by(item_name) %>%
  mutate(Pchange3 = value / lag(value, 1) - 1) %>%
  ungroup() %>%
  group_by(date) %>%
  mutate(
    normalized = sum(weight, na.rm = TRUE),
    weightN    = weight / normalized
  ) %>%
  arrange(Pchange3) %>%
  mutate(
    cumsum  = cumsum(weight) / 100,
    cumsumN = cumsum(weightN)
  ) %>%
  ungroup() %>%
  mutate(Pchange3a = (1 + Pchange3)^12 - 1)
```

---

## Plot

```{r plot, fig.width=11, fig.height=9}
median_df %>%
  mutate(dateF = as.factor(date)) %>%
  filter(cumsumN <= trim_upper, cumsumN >= trim_lower) %>%
  filter(date >= as.Date(chart_start_date)) %>%
  filter(month(date) %in% quarters) %>%
  mutate(
    monthC  = format(date, "%B, %Y"),
    monthC  = fct_reorder(monthC, date),
    monthCR = fct_rev(monthC)
  ) %>%
  ggplot(aes(x = Pchange3a, y = monthCR, fill = after_stat(x))) +
  geom_density_ridges_gradient() +
  scale_fill_viridis(option = "H") +
  theme_ridges() +
  theme(legend.position = "none") +
  scale_x_continuous(labels = percent) +
  labs(
    title    = "CPI Inflation Report: Distribution Ridgeline Plot",
    subtitle = paste0(
      "Trimmed distribution: dropping bottom ", trim_lower * 100,
      "% and top ", (1 - trim_upper) * 100, "% by weight"
    ),
    x       = "One Annualized Month Percent Change",
    y       = "",
    caption = "OER is treated as one value. h/t Mike Konczal"
  ) +
  theme(
    plot.title.position = "plot",
    plot.title    = element_text(size = 20, margin = margin(0, 0, 5, 0)),
    plot.subtitle = element_text(size = 13),
    plot.caption  = element_text(size = 10, face = "italic"),
    axis.text.y   = element_text(size = 12, face = "bold"),
    axis.text.x   = element_text(size = 12)
  )
```
